---
title: "Spatial Data"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE)
```

This document introduces various R packages for representing spatial data in R. It also explores packages for reading/writing spatial data from external formats.

# Vecor Data Using `sp` Package

## Construct a `sp` object.

Read the `meuse` data.frame and convert it into a spatial object. Explore the result.

```{r sp-01}
library(sp)
data(meuse)
print(class(meuse))
print(colnames(meuse))
{{ coordinates(meuse) <- ~x+y }}
{{ proj4string(meuse) <- CRS("+init=epsg:28992") }}
print(class(meuse))
summary(meuse)
```

## Load `sp` objects from packages.

Load all 'meuse' datasets

```{r sp-02}
demo("meuse", ask=FALSE)
```

## Read `sp` objects from files

```{r sp-03}
library(rgdal)
dsn <- system.file("vectors", package = "rgdal")[1]

ogrInfo(dsn=dsn, layer="scot_BNG")

print(OGRSpatialRef(dsn=dsn, layer="scot_BNG"))

scot_BNG <- readOGR(dsn=dsn, layer="scot_BNG")

print(class(scot_BNG))

summary(scot_BNG)
```

## Selecting/Subsetting

```{r sp-04}
# 9th and 10th rows, and lead and zinc columns
meuse[9:10, c('lead','zinc')]
# Filtering on a criteria
dim(meuse[meuse$lead>quantile(meuse$lead,.75),])

# Attribute data
range(meuse[['zinc']])
```

## Projection Transformation

```{r sp-05}
print(proj4string(meuse))
meuse.latlon <- spTransform(meuse, CRS("+init=epsg:4326")) # What's '+init=epsg:4326' ?
print(proj4string(meuse))
```

## Other Operations

- Vignettes [1](https://cran.r-project.org/web/packages/sp/vignettes/intro_sp.pdf) and [2](https://cran.r-project.org/web/packages/sp/vignettes/over.pdf) for `sp`.
- Packages [gdistance](https://cran.r-project.org/web/packages/gdistance), [geosphere](https://cran.r-project.org/web/packages/geosphere).

# Vector Data Using `sf` Package

`sf` is the successor to `sp` and provides much faster and `tidyverse` compatible API.

## Read a `sf` Object

```{r sf-01}
library(sf)
scot_BNG <- sf::st_read(system.file("vectors", package = "rgdal")[1], 'scot_BNG')
print(class(scot_BNG))
summary(scot_BNG)
```

## `sf` operations

```{r sf-02}
# Just the geometry
scot_BNG.geom <- st_geometry(scot_BNG)
print(class(scot_BNG.geom))

# Reprojection
st_crs(scot_BNG)
scot_BNG.latlon <- st_transform(scot_BNG, '+init=epsg:4326')
st_crs(scot_BNG.latlon)

# Centroids
summary(scot_BNG$geometry)
scot_BNG.centers <- st_centroid(scot_BNG)
summary(scot_BNG.centers$geometry)
```

## Other Operations

```{r sf-03}
methods(class='sf')
```

# Raster Data using `raster` Package

## Read a Raster file

```{r raster-01}
library(raster)
r <- raster(system.file("external/test.grd", package="raster"))
print(class(r))
summary(r)
```

## `raster` Operations

```{r raster-02, fig.align='center', fig.width=12,fig.height=6}
r_log <- calc(r, log)
par(mar=c(5,5,5,5), mfrow=c(1,2))
hist(r); hist(r_log)
```

For many more see the `raster` package [vignette](https://cran.r-project.org/web/packages/raster/vignettes/Raster.pdf).
