---
title: "Geospatial Visualization using R"
subtitle: "Part 1: Introduction"
author: "Bhaskar V. Karambelkar"
date: "2017/07/04"
output:
  xaringan::moon_reader:
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: ["default", "xaringan-style.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
        click: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE)
library(magrittr)
```

# Part 1: Welcome! / Bienvenue! / Welgekomen!

- What's this all about?

- The Tutor/Presenter

- The Audience 

- Scope of the Tutorial

- Setup

- On to Part 2

---

# The Tutor/Presenter

- Day job is not GIS related, but it's a hobby that I take very seriously.

- Works in Information Security. Don't you ever "CYBER" me!

- Loves maps and loves R.

- Had a boss who loves maps and R even more. Bob Rudis!

- Contributor to the 'leaflet' package, and author of `leaflet.extras`/`leaflet.esri`/`tilegramsR`/`widgetframe` packages.

- Now your turn!

---

# Obligatory Hello World! Get it?

```{r 01, fig.width=10, fig.height=6, fig.align='center'}
maps::map('world', fill=TRUE, col='#ffA50077', border='black', bg='#00000011')
```

---

# Tutorial Scope

- Part 1: Welcome! .gray[[9:30 - 9:50]]

- Part 2: R Packages .gray[[9:50 - 10:05]]

- Part 3: Spatial Data .gray[[10:05 - 10:20]]

- Part 4: Spatial Data Operations .gray[[10:25 - 11:40]]

- Part 5: Ways to Output Maps .gray[[10:40 - 11:00]]

- Part 6: Static Maps .gray[[11:30 - 12:05]]

- Part 7: Interactive Maps .gray[[12:10 - 12:50]]

- Part 8: Q/A .gray[[12:50 - 1:00]]

---

# Getting Setup

- [Docker](https://store.docker.com/search?type=edition&offering=community) FTW!

- [Rocker](https://hub.docker.com/u/rocker/) FTW!

- `docker pull bhaskarvk/rgeodataviz`

- `docker run -d -p 8787:8787 --name rgeodataviz \` <br/>`bhaskarvk/rgeodataviz`

- Go to http://localhost:8787. You should see RStudio. Login rstudio/rstudio. 

- You should see `user2017.geodataviz` in the folders view, click on it and then on the `user2017.geodataviz.Rproj` file. 'Yes' on the confirmation box.

- If docker is not an option, see [this](https://gist.github.com/bhaskarvk/32b2a03a430a04a67664b85cad89c825) gist for details of all the packages needed.

---

class: inverse middle

# Part 1: The End! 

Continue to [Part 2: R Packages](02-R_Packages.html)

.footnote[Restart [Part1](01-Introduction.html)]
