---
title: "Geospatial Visualization using R"
subtitle: "Part 4: Spatial Data Operations"
author: "Bhaskar V. Karambelkar"
date: "2017/07/04"
output:
  xaringan::moon_reader:
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: ["default", "xaringan-style.css"]
    nature:
      highlightStyle: zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
        click: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'svg')
options(htmltools.dir.version = FALSE)
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
knitr::opts_chunk$set(cache = TRUE)
library(magrittr)
```

# Selectiing features/fields `sp` objects

```{r 04-01, echo=FALSE, include=FALSE}
library(sp)
data(meuse)
coordinates(meuse) <- ~x+y
proj4string(meuse) <- CRS("+init=epsg:28992")
scot_BNG <- rgdal::readOGR(
  dsn = system.file("vectors", package = "rgdal")[1],
  layer = "scot_BNG")
```

```{r 04-02}
dim(meuse@data)
colnames(meuse@data)
print(meuse[9:10, c('lead','zinc')])
nrow(meuse[meuse$lead>quantile(meuse$lead,.75),])
```

---
# Other Operations on `sp` objects

--

- `[[`: Data for single column `plot(hist(meuse[['zinc']]))`

--

- `[[<-`: Replace values of single column <br/> `meuse[['lead']] <- rnorm(nrow(meuse))`

--

- `proj4String` / `proj4String<-`: Returns / sets (w/o transforming) the CRS.

--

- `spTransform(meuse, CRS("+init=epsg:4326"))`: Convert from one CRS to another.

--

- `geometry(meuse)`: Strip the fields and return the bare geometry.

--

- `over(points, polygons, returnList=TRUE)`:<br/> Returns a list indicating which polygons contain each point in `points` data-set.

--

- `rgeos` / `gdistance` / `geosphere` / `maptools`:<br/>Packages contain many geometry and feature related operations.

---

# Example

```{r 04-03, fig.height=5, fig.align='center'}
par(mar=c(0,0,0,0))
plot(scot_BNG, col='#aaaaaa', border=0)
{{plot(rgeos::gCentroid(scot_BNG, byid=TRUE), add=TRUE, pch=20) }}
```

---

# Operations on `sf` Objects

--

- `sp` => `sf`

```{r 04-04}
suppressPackageStartupMessages(library(sf))
{{meuse_sf <- st_as_sf(meuse) }}
as.data.frame(meuse_sf[1:2,c('lead', 'zinc', 'geometry')])
```

--

- `st_coordinates(sf)`: Extract the X/Y coordinates.

--

- `st_geometry(sf)`: Just the geometry, the `sfc` column.

--

- `st_as_text(sfg)` / `st_as_binary(sfg)` convert to WKT and WKB resp.

- `st_as_sfc()` for converting a foreign object into a `sfc`.

--

- Check out `methods(class='sfc')` for many more methods.<sup>1</sup>

--

.footnote[1 - Best place to start are the `sf` vignettes [here](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html) & [here](https://cran.r-project.org/web/packages/sf/vignettes/sf3.html).]

---

# Example

```{r 04-05, fig.height=4, fig.align='center'}
{{usa_sfg <- st_transform(st_geometry(albersusa::usa_sf()), albersusa::us_aeqd_proj) }}
{{usa_c <- st_centroid(usa_sfg) }}
{{usa_50 <- (usa_sfg - usa_c) * .50 + usa_c }}
par(mar=c(0,0,0,0));plot(usa_sfg)
plot(usa_50, add=TRUE, col='#FFA50077', border=0)
plot(usa_c, add=TRUE, pch=20)
```

---

# Operations on `raster` Objects

```{r 04-06}
suppressPackageStartupMessages(library(raster))
r <- raster(system.file("external/test.grd", package="raster"))
inMemory(r)
nlayers(r)
dim(r)
cellStats(r, function(x, ...) 
  c(min=min(x,...), mean=mean(x,...), max=max(x,...)))
```

---

# Operations on `raster` Objects

```{r 04-07, fig.align='center', fig.width=12,fig.height=4.5}
{{ r_log <- calc(r, log) }}
par(mar=c(5,5,5,5), mfrow=c(1,2))
hist(r); hist(r_log)
```


---

# Operations via External Tools

--

- `RQGIS`: Uses QGIS Python API and allows you to access many native and 3<sup>rd</sup>-party plugin based geospatial algorithms/analysis.

--

- `RQGIS` also allows you to access `GRASS7` and `SAGA`, but you also have option of using `rgrass7` and `RSAGA`.

--

- `rpostgis` & `postGIStools` allow you to access data in PostGIS, but so does the PostGIS driver in `gdal` so you can query directly from `rgdal` and `sf` too.

---

class: inverse middle

# Part 4: The End! 

Continue to [Part 5: Ways to Output Maps](05-Ways_to_Output_Maps.html)

.footnote[Restart [Part 4: Spatial Data Operations](04-Spatial_Data_Operations.html)]
