---
title: "Geospatial Visualization using R"
subtitle: "Part 6: Static Maps"
author: "Bhaskar V. Karambelkar"
date: "2017/07/04"
output:
  xaringan::moon_reader:
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: ["default", "xaringan-style.css"]
    nature:
      highlightStyle: zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
        click: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'svg')
options(htmltools.dir.version = FALSE)
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
knitr::opts_chunk$set(cache = TRUE)
```

# Ways To Create Static Maps

--

- Base plotting

--

- `ggplot2` and related packages

--

- Ease-of-use packages

--

  - `tmap`
  - `choroplethr`
  - `cartography`
  - `ggmap`

---

# Plotting with Base R: Sample 1

--

- `plot()` works on all spatial objects. `plot(..., add=TRUE)` to overlay.

--

- Can be cumbersome to create production-ready maps.

--

```{r 06-00}
suppressPackageStartupMessages(library(sp))
demo(meuse, ask = FALSE, echo = FALSE) 
crs.longlat = CRS("+init=epsg:4326")
meuse.longlat = spTransform(meuse, crs.longlat)
meuse.riv.longlat <- spTransform(meuse.riv, crs.longlat)
grid.lines <- gridlines(meuse.longlat)
```

```{r 06-01, eval=FALSE}
par(mar = c(1, 1, 1, 1))
plot(methods::as(meuse.longlat, "Spatial"), expandBB=c(0.05,0,0.1,0))
plot(grid.lines, add = TRUE, col = grey(.8))
plot(meuse.longlat, pch=1, cex = sqrt(meuse$zinc)/12, add = TRUE)
text(labels(grid.lines, side=2:3), col = grey(.7), offset=1.5)
v = c(100,200,400,800,1600)
legend("bottomright", legend = v, pch = 1, pt.cex = sqrt(v)/12,
       text.col =grey(.8), box.col=grey(0.8), title='Zinc Conc. (ppm)')
plot(meuse.riv.longlat, add = TRUE, col = grey(.9, alpha = .5))
```

---

# Plotting with Base R: Sample 1

```{r 06-02, echo=FALSE,  fig.width=6.5, fig.height=5.5, fig.align='center'}
par(mar = c(1, 1, 1, 1))
plot(as(meuse.longlat, "Spatial"), expandBB=c(0.05,0,0.1,0))
plot(grid.lines, add = TRUE, col = grey(.8))
plot(meuse.longlat, pch=1, cex = sqrt(meuse$zinc)/12, add = TRUE)
text(labels(grid.lines, side=2:3), col = grey(.7), offset=1.5)
v = c(100,200,400,800,1600)
legend("bottomright", legend = v, pch = 1, pt.cex = sqrt(v)/12,
       text.col =grey(.8), box.col=grey(0.8), title='Zinc Conc. (ppm)')
plot(meuse.riv.longlat, add = TRUE, col = grey(.9, alpha = .5))
```

---

# Plotting with Base R: Sample 2

```{r 06-03, eval=FALSE}
library(maps)
data(world.cities)
world.cities <- world.cities[world.cities$pop>1000000,]
coordinates(world.cities) <- ~long+lat 
proj4string(world.cities) <- '+init=epsg:4326'

world <- rnaturalearth::countries110
world <- world[world$name != 'Antarctica',]

grid.lines.mj <- gridlines(world,easts = seq(-180,180,by=30), norths = seq(-90,90,by=30))
grid.lines.mi <- gridlines(world,easts = seq(-165,195,by=15), norths = seq(-90,90,by=15))

par(mar = c(8, 0.1, 0.1, 0.1))
plot(methods::as(world, 'Spatial'), expandBB=c(0,0,0.05,0.05))
plot(world, add=TRUE, border=grey(0.2))
plot(grid.lines.mi, col=grey(0.95), add=T)
plot(grid.lines.mj, col=grey(0.9), add=T)
text(labels(grid.lines.mj, side=1:2, col = grey(.6), offset=0.3)
plot(world.cities, add=TRUE, col='#FF5A0088', pch=20,
     cex=world.cities$pop/2000000)
v = c(1,4,8,12)
legend("topright", legend = v, pch = 20, pt.cex = v/2,
       text.col =grey(.7), box.col=grey(0.9), col = '#FF5A0088',
       title='Pop. (Millions)', horiz =T)
```

---

# Plotting with Base R: Sample 2

```{r 06-04, echo=FALSE,  fig.width=10, fig.height=7, fig.align='center'}
library(maps)
data(world.cities)
world.cities <- world.cities[world.cities$pop>1000000,]
coordinates(world.cities) <- ~long+lat 
proj4string(world.cities) <- '+init=epsg:4326'

world <- rnaturalearth::countries110
world <- world[world$name != 'Antarctica',]

grid.lines.mj <- gridlines(world,easts = seq(-180,180,by=30), norths = seq(-90,90,by=30))
grid.lines.mi <- gridlines(world,easts = seq(-165,195,by=15), norths = seq(-90,90,by=15))
world.cities <- spTransform(world.cities, CRS("+proj=wintri"))
world <- spTransform(world, CRS("+proj=wintri"))
grid.lines.mj <- spTransform(grid.lines.mj,CRS("+proj=wintri"))
grid.lines.mi <- spTransform(grid.lines.mi,CRS("+proj=wintri"))

par(mar = c(8, 0.1, 0.1, 0.1))
plot(methods::as(world, 'Spatial'), expandBB=c(0,0,0.05,0.05))

plot(grid.lines.mi, col=grey(0.95), add=T)
plot(grid.lines.mj, col=grey(0.9), add=T)
text(labels(grid.lines.mj, side=1:2, labelCRS = CRS("+init=epsg:4326")), col = grey(.6), offset=0.3)

plot(world, add=TRUE, border=grey(0.2), col=grey(0.9))
plot(world.cities, add=TRUE, col='#FF5A0088', pch=20,
     cex=world.cities$pop/2000000)

v = c(1,4,8,12)
legend("topright", legend = v, pch = 20, pt.cex = v/2,
       text.col =grey(.7), box.col=grey(0.9),
       col = '#FF5A0088',
       title='Pop. (Millions)', horiz =T)
```

---

# Plotting with Base R: Sample 2

--

- Actually I cheated with the last plot. 

--

- The plot you saw as plotted in [Winkel triple](https://en.wikipedia.org/wiki/Winkel_tripel_projection) projection, while the code was using lat/long values.

--

- Before the `par(...)` line we need to convert all the data to our target projection.

```{r 06-05, eval=FALSE}
world <- spTransform(world, CRS("+proj=wintri"))
world.cities <- spTransform(world.cities, CRS("+proj=wintri"))
grid.lines.mj <- spTransform(grid.lines.mj,CRS("+proj=wintri"))
grid.lines.mi <- spTransform(grid.lines.mi,CRS("+proj=wintri"))

```

--

- But the lables were showing lat/long values, how come ?

--

```{r 06-06, eval=FALSE}
text(labels(grid.lines.mj, side=1:2, labelCRS = CRS("+init=epsg:4326")),
     col = grey(.6), offset=0.3)
```

--

- [projectionwizard.org](http://projectionwizard.org/) helps you determine sensible projection/s for your data.

---

# Plotting with Base R: Miscellaneous Notes


- `sp::gridlines()`, `sp::graiddata()`, `sp::labels()`, `sp::text()`: <br/> Creates and plot EW and NS grid lines

--

- `sp::bbox()` / `sp::bbexpand()`: Retrieves / expands the bounding-box of a spatial object.

--

- Raster data is plotted similar to vector data, but for reprojecting rasters use `raster::projectRaster()` instead of `sp::Transform()`.

--

- [`sp` Gallery](https://edzer.github.io/sp/) is probably the best source for advanced `sp` plotting using base R graphics. 

--

- [`sf` Plotting](http://r-spatial.org/r/2017/01/12/newssf.html) is discussed in detail by Edzer Pebesma, the author of `sf`.

--

- Robin Lovelace maintains a really good [repository](https://github.com/Robinlovelace/Creating-maps-in-R) intruducing spatial visualization in R.

--

- Zev Ross (again!) has a very good [blog post](http://zevross.com/blog/2015/03/30/map-and-analyze-raster-data-in-r/) for working with raster data in R.

---

# Plotting with `ggplot2`: `sp` Data

```{r 06-07}
world <- rnaturalearth::countries110
europe <- world[world$region_un=="Europe"&world$name!='Russia',]
# plot(europe)

# Let's add a unique ID column to our data.
{{europe@data$id <- row.names(europe@data)}}

# A bounding box for continental Europe.
europe.bbox <- SpatialPolygons(list(Polygons(list(Polygon(
  matrix(c(-25,29,45,29,45,75,-25,75,-25,29),byrow = T,ncol = 2)
)), ID = 1)), proj4string = CRS(proj4string(europe)))

# Get polygons that are only in continental Europe.
europe.clipped <-
{{  rgeos::gIntersection(europe, europe.bbox, byid = TRUE, id=europe$id)}}

```

- `ggplot2` requires tidy data, which `sp` objects are not.

```{r 06-08}
europe.tidy <- broom::tidy(europe.clipped)
europe.tidy <- dplyr::left_join(europe.tidy, europe@data, by='id')
```

---

# Plotting with `ggplot2`: `sp` Data

- Ready to make a production grade map?

--

```{r 06-09, eval=FALSE}
library(ggplot2)
ggplot(europe.tidy, aes(long,lat, group=group,fill=gdp_md_est/1000)) +
  geom_polygon(alpha=0.8,color='black') +
  coord_map("azequalarea") +
  hrbrthemes::theme_ipsum_rc() +
  viridis::scale_fill_viridis(
    name='Median GDP \n(in Billions)', direction = -1, labels=scales::dollar) +
  labs(x=NULL, y=NULL, 
       title='Median GDP Estimates of\nContinental Europe & Iceland',
       caption='Source: http://www.naturalearthdata.com/')

```


---

# Plotting with `ggplot2`: `sp` Data

```{r 06-10, echo=FALSE,  fig.width=10, fig.height=6, fig.align='center'}
library(ggplot2)
ggplot(europe.tidy, aes(long,lat, group=group,fill=gdp_md_est/1000)) +
  geom_polygon(alpha=0.8,color='black') +
  coord_map("azequalarea") +
  hrbrthemes::theme_ipsum_rc() +
  viridis::scale_fill_viridis(
    name='Median GDP \n(in Billions)', direction = -1, labels=scales::dollar) +
  labs(x=NULL, y=NULL, 
       title='Median GDP Estimates of\nContinental Europe & Iceland',
       caption='Source: http://www.naturalearthdata.com/')

```

---

# Plotting with `ggplot2`: `sf` Data

- Similar to last plot, but with `sf`.

--

```{r 06-11, eval=FALSE}

library(sf)

world <- st_as_sf(rnaturalearth::countries110)
europe <- dplyr::filter(world, region_un=="Europe" & name!='Russia')

# A bounding box for continental Europe.
europe.bbox <- st_polygon(list(
  matrix(c(-25,29,45,29,45,75,-25,75,-25,29),byrow = T,ncol = 2)))

europe.clipped <- st_intersection(europe, st_sfc(europe.bbox, crs=st_crs(europe)))

ggplot(europe.clipped, aes(fill=gdp_md_est/1000)) +
  geom_sf(alpha=0.8,col='white') +
   coord_sf(crs="+proj=aea +lat_1=36.333333333333336 +lat_2=65.66666666666667 +lon_0=14") +
  hrbrthemes::theme_ipsum_rc() +
  viridis::scale_fill_viridis(
    name='Median GDP \n(in Billions)', direction = -1, labels=scales::dollar) +
  labs(x=NULL, y=NULL, title=NULL,
       caption='Source: http://www.naturalearthdata.com/')
```

---

# Plotting with `ggplot2`: `sf` Data

```{r 06-12, echo=FALSE,  fig.width=10, fig.height=6, fig.align='center'}
suppressPackageStartupMessages(library(sf))

world <- st_as_sf(rnaturalearth::countries110)
europe <- dplyr::filter(world, region_un=="Europe" & name!='Russia')

# A bounding box for continental Europe.
europe.bbox <- st_polygon(list(
  matrix(c(-25,29,45,29,45,75,-25,75,-25,29),byrow = T,ncol = 2)))

europe.clipped <- suppressWarnings(st_intersection(europe, st_sfc(europe.bbox, crs=st_crs(europe))))


ggplot(europe.clipped, aes(fill=gdp_md_est/1000)) +
  geom_sf(alpha=0.8,col='white') +
  coord_sf(crs="+proj=aea +lat_1=36.333333333333336 +lat_2=65.66666666666667 +lon_0=14") +
  hrbrthemes::theme_ipsum_rc() +
  viridis::scale_fill_viridis(
    name='Median GDP \n(in Billions)', direction = -1, labels=scales::dollar) +
  labs(x=NULL, y=NULL, title=NULL,
       caption='Source: http://www.naturalearthdata.com/')
```

---

# Plotting with `ggplot2`:`raster` Data

```{r 06-13, fig.width=6, fig.height=4, fig.align='center'}
suppressPackageStartupMessages(library(raster))
#suppressPackageStartupMessages(library(rasterVis))
r <- raster(system.file("external/test.grd", package="raster"))

{{rasterVis::gplot(r) + geom_tile(aes(fill = value)) }} +
  viridis::scale_fill_viridis(direction = -1, na.value='#FFFFFF00') +
  coord_equal() + hrbrthemes::theme_ipsum()
```

---

# Plotting with `ggplot2`: Notes

- `ggplot2` is huge! `geoms`/`scales`/`legends`/`themes`/`facets`...

- There are several ggplot2 [extension](https://www.ggplot2-exts.org) packages.

- Prefer `geom_polygon` over `geom_map`.

- Be careful when tidying and subsequently joining `sp` data.

- Add the main dataset to `ggplot2()`, and others to respective `geom`'s `data` argument. You may need `inherit.aes=F` in `geom`s with their own data.

- Some good map themes `ggthemes::theme_map`, `hrbrthemes::theme_ipsum`, `ggplot2::theme_bw`.

- For `sp` objects prefer `ggalt::coord_proj` over `ggplot2::coord_map`, as the former supports proj4 strings.

- `sf` support is not on CRAN and evolving, expect bugs/issues.

---

# Plotting with tmap

- `plot` mode for static maps and `view` mode for interactive maps.

- `qtm()` Quick plotting. Otherwise `tm_shape()` + `tm_<feature>`.

- Syntax similar to `ggplot2`, but returned object is not a `ggplot2` plot.

- Unlike `ggplot2` you need one `tm_shape` + one `tm_<feature>` per layer.

- Supports many different map styles and aesthetic elements like compass, legend, scale etc.

--

```{r 06-14, eval=FALSE}
usa_pop_history <- readr::read_tsv(system.file('extdata','usa_pop_history.tsv', package='user2017.geodataviz'))
usa <- dplyr::left_join(albersusa::usa_sf(), usa_pop_history, by=c('name'='State'))
usa <- st_transform(usa, crs=albersusa::us_laea_proj)

years <-  c(1900,1950,1990,2000,2010,2015)

tmap::qtm(usa, fill=paste0("p.",years), title=years, 
          layout.title.color='red', layout.title.position=c('center','top'),
          layout.main.title = 'U.S.A. Population by State',
          style = 'col_blind', facets.nrow=2)
```

---

# Plotting with `tmap`

```{r 06-15, echo=FALSE, fig.width=12, fig.height=6, fig.align='center'}
usa_pop_history <- suppressMessages(
  readr::read_tsv(system.file('extdata','usa_pop_history.tsv', package='user2017.geodataviz')))
usa <- suppressWarnings(dplyr::left_join(albersusa::usa_sf(), usa_pop_history, by=c('name'='State')))
usa <- st_transform(usa, crs=albersusa::us_laea_proj)
years <-  c(1900,1950,1990,2000,2010,2015)
tmap::qtm(usa, fill=paste0("p.",years), title=years, 
          layout.title.color='red',  layout.title.position=c('center','top'),
          layout.main.title = 'U.S.A. Population by State',
          style = 'col_blind', facets.nrow=2)
```

---

# Plotting with `ggmap`

--

- Primarily used to add base maps from Google Maps, OSM, Stamen etc. to `ggplot2` maps.

--

- The returned object is a `ggplot2` object which can be further worked upon.

--

```{r 06-16, eval=FALSE}
library(ggplot2)
library(ggmap)

paste0("2016-0",1:7) %>%
  purrr::map(function(month) {
    suppressMessages(readr::read_csv(
      system.file(
        sprintf("examples/data/London-Crimes/%s/%s-city-of-london-street.csv.zip",
                month,month), package='leaflet.extras')
    ))
  }) %>%
  dplyr::bind_rows() -> crimes

qmplot(Longitude, Latitude, data=crimes, geom="blank", zoom=15,
       maptype = "toner-lite", facets = ~Month) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", alpha = .3) +
  scale_fill_gradient2("Crime Heatmap", low = "white", mid = "yellow", high = "red") + theme(legend.position = 'hide')
```

---

# Plotting with `ggmap`

```{r 06-17, echo=FALSE,  fig.width=10, fig.height=6, fig.align='center'}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggmap))

paste0("2016-0",1:7) %>%
  purrr::map(function(month) {
    suppressMessages(readr::read_csv(
      system.file(
        sprintf("examples/data/London-Crimes/%s/%s-city-of-london-street.csv.zip",
                month,month),
        package='leaflet.extras')
    ))
  }) %>%
  dplyr::bind_rows() -> crimes

suppressMessages(suppressWarnings(
  qmplot(Longitude, Latitude,
         data = crimes %>% dplyr::filter(!is.na(Latitude)),
         geom="blank", zoom=15,
         maptype = "toner-lite", facets = ~Month) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", alpha = .3) +
    scale_fill_gradient2("Crime Heatmap", low = "white", mid = "yellow", high = "red")  + theme(legend.position = 'hide')
))
```
---

# Plotting with Ease-of-Use Pkgs: cartography

---

# Plotting with Ease-of-Use Pkgs: cartography

---


# Plotting with Ease-of-Use Pkgs: choroplethr

---

# Plotting with Ease-of-Use Pkgs: chorophethr

---

class: inverse middle

# Part 5: The End!

Continue to [Part 7: Dynamic Maps](07-Dynamic Maps.html)

.footnote[Restart [Part 6: Static Maps](05-Static Maps.html)]
