---
title: "Geospatial Visualization using R"
subtitle: "Part 7: Interactive Maps"
author: "Bhaskar V. Karambelkar"
date: "2017/07/04"
output:
  xaringan::moon_reader:
    lib_dir: libs
    chakra: libs/remark-latest.min.js
    css: ["default", "xaringan-style.css"]
    nature:
      highlightStyle: zenburn
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
        click: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'svg')
options(htmltools.dir.version = FALSE)
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
knitr::opts_chunk$set(cache = TRUE)
library(magrittr)
```

# `ggiraph` for simple effects

- Use `geom_*_interactive` with hover, tooltip, onClick aesthetics, plot with `ggiraph`.

```{r 07-01, eval=FALSE}
library(ggplot2); suppressPackageStartupMessages(library(ggiraph))

africa.spdf <- methods::as(africa, 'Spatial')
africa.spdf@data$id <- row.names(africa.spdf@data)

africa.tidy <- broom::tidy(africa.spdf)
africa.tidy <- dplyr::left_join(africa.tidy, africa.spdf@data, by='id')

g <- ggplot(africa.tidy) +
  geom_polygon_interactive(
    color='black',
    aes(long, lat, group=group, fill=internet.usage.2015,
        tooltip=sprintf("%s<br/>%s",iso_a3,internet.usage.2015))) +
 ggthemes::theme_map() +
  colormap::scale_fill_colormap(
    colormap=colormap::colormaps$copper, reverse = T) +
  labs(title='Internet Usage in Africa in 2015', subtitle='As Percent of Population',
       caption='Source: World Bank Open Data.')

ggiraph(code=print(g))
```

---

# `ggiraph` plot

```{r 07-02, include=FALSE}
library(sp)
library(sf)
library(ggplot2)
library(ggiraph)
```

```{r 07-03, echo=FALSE, fig.width=10, fig.height=8, fig.align='center'}
world <- sf::st_as_sf(rnaturalearth::countries110)
internet_usage <- suppressMessages(readr::read_csv(system.file(
  'extdata','africa-internet_usage-2015.csv', package='user2017.geodataviz')))

africa <- dplyr::filter(world, region_un=='Africa') %>%
  dplyr::left_join(internet_usage %>% dplyr::select(
    `Country Code`, `2015 [YR2015]`
  ) %>% dplyr::rename(iso_a3=`Country Code`, internet.usage.2015=`2015 [YR2015]`),
  by = 'iso_a3') %>%
  st_transform(crs="+proj=laea +lon_0=18.984375")

africa.spdf <- methods::as(africa, 'Spatial')
africa.spdf@data$id <- row.names(africa.spdf@data)

africa.tidy <- suppressMessages(broom::tidy(africa.spdf))
africa.tidy <- dplyr::left_join(africa.tidy, africa.spdf@data, by='id')

g <- ggplot(africa.tidy) +
  geom_polygon_interactive(
    color='black',
    aes(long, lat, group=group, fill=internet.usage.2015,
        tooltip=sprintf("%s<br/>%s",iso_a3,internet.usage.2015))) +
 ggthemes::theme_map() +
  colormap::scale_fill_colormap(
    colormap=colormap::colormaps$copper, reverse = T) +
  labs(title='Internet Usage in Africa in 2015', subtitle='As Percent of Population',
       caption='Source: World Bank Open Data.')

widgetframe::frameWidget(ggiraph(code=print(g)))
```

---

# `plotly`  

- `htmlwidgets` based wrapper for [plot.ly](https://plot.ly/) JavaScript library.

- Extremely easy and yet powerful. Commercial use requires a paid version.

- Plays nicely with `ggplot2`

```{r 07-04, eval=FALSE}
library(plotly)
usa <- albersusa::usa_sf("laea")
usd <- crosstalk::SharedData$new(usa)
p <- ggplot(usd) + geom_sf(aes(fill = pop_2010))

ggplotly(p) %>%
  highlight(
    "plotly_hover",
    selected = attrs_selected(line = list(color = "black"))
)

```

---

# `plotly` plot

```{r 07-041, echo=FALSE, fig.width=10, fig.height=8, fig.align='center'}
suppressPackageStartupMessages(library(plotly))
usa <- albersusa::usa_sf("laea")
usd <- crosstalk::SharedData$new(usa)
p <- ggplot(usd) + geom_sf(aes(fill = pop_2010))

ggplotly(p) %>%
  highlight(
    "plotly_hover",
    selected = attrs_selected(line = list(color = "black"))
) %>%
widgetframe::frameWidget()
```

---

# `rBokeh`

- Port of Python's [bokeh](http://bokeh.pydata.org/en/latest/) library.

- Has its own syntax.

```{r 07-05, eval=FALSE}
library(maps)
data(world.cities)
library(rbokeh)
caps <- dplyr::filter(world.cities, capital == 1)
caps$population <- prettyNum(caps$pop, big.mark = ",")
suppressWarnings(figure(width = 800, height = 450, padding_factor = 0) %>%
  ly_map("world", col = "gray") %>%
  ly_points(long, lat, data = caps, size = 5,
            hover = c(name, country.etc, population)))

```

---

# `rBokeh` plot

```{r 07-06, echo=FALSE, fig.width=10, fig.height=8, fig.align='center'}

library(maps)
data(world.cities)
library(rbokeh)
caps <- dplyr::filter(world.cities, capital == 1)
caps$population <- prettyNum(caps$pop, big.mark = ",")
plot <- suppressWarnings(figure(width = 800, height = 450, padding_factor = 0) %>%
  ly_map("world", col = "gray") %>%
  ly_points(long, lat, data = caps, size = 5,
            hover = c(name, country.etc, population)))

widgetframe::frameWidget(plot, width=1024, height=576)
```

---

# `highcharter`

- Wraps [highcharts](https://www.highcharts.com/), a very popular JavaScript library.

```{r 07-07, eval=FALSE}

download_map_data <- memoise::memoise(highcharter::download_map_data)
get_data_from_map <- memoise::memoise(highcharter::get_data_from_map)
hcmap <- memoise::memoise(highcharter::hcmap)

mapdata <- get_data_from_map(download_map_data("countries/au/au-all"))

data_fake <- mapdata %>%
  dplyr::select(code = `hc-a2`) %>%
  dplyr::mutate(value = 1e5 * abs(rt(nrow(.), df = 10)))

hcmap("countries/au/au-all", data = data_fake, value = "value",
        joinBy = c("hc-a2", "code"), name = "Fake data",
        dataLabels = list(enabled = TRUE, format = '{point.name}'),
        borderColor = "#FAFAFA", borderWidth = 0.1,
        tooltip = list(valueDecimals = 2, valuePrefix = "$", valueSuffix = " AUD")) %>%
  highcharter::hc_title(text="Economy Down Under")

```

---

`highcharter` plot

```{r 07-08, echo=FALSE, fig.width=10, fig.height=8, fig.align='center'}
library(magrittr)

download_map_data <- memoise::memoise(highcharter::download_map_data)
get_data_from_map <- memoise::memoise(highcharter::get_data_from_map)
hcmap <- memoise::memoise(highcharter::hcmap)

mapdata <- get_data_from_map(download_map_data("countries/au/au-all"))

data_fake <- mapdata %>%
  dplyr::select(code = `hc-a2`) %>%
  dplyr::mutate(value = 1e5 * abs(rt(nrow(.), df = 10)))

hcmap("countries/au/au-all", data = data_fake, value = "value",
        joinBy = c("hc-a2", "code"), name = "Fake data",
        dataLabels = list(enabled = TRUE, format = '{point.name}'),
        borderColor = "#FAFAFA", borderWidth = 0.1,
        tooltip = list(valueDecimals = 2, valuePrefix = "$", valueSuffix = " AUD")) %>%
  highcharter::hc_title(text="Economy Down Under") %>%
  widgetframe::frameWidget()

```

---

# `leaflet`

```{r libs, echo=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(widgetframe))
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(leaflet.extras))
```

- Mother of all interactive mappings!

- Extremely powerful and Extremely flexible.

- Current State

  - `leaflet`: Developed and actively maintained by RStudio.
  
  - `leaflet.extras`, `leaflet.esri`: Developed and maintained by yours truly. Provides additional plotting options using leaflet plugins.
  
  - `mapview`, `mapedit`: Developed and maintained by the good guys at [r-spatial](http://r-spatial.org/projects/). Allow interactive exploratory spatial analysis.

- There is another unrelated project `leafletR` [CRAN Page](https://github.com/chgrl/leafletR), but really don't know much about it.

---

# `leaflet` One Page Summary

```{r 07-leaflet-00, eval=FALSE}
leaflet(data) | leafletProxy() %>% 
  
  
  setView(lat, lon, zoom) # Initial View OR
  fitBounds(lat_se, lon_se, latnw, lon_nw) # Initial Bounds
  setMaxBounds(lat_se, lon_se, latnw, lon_nw) # Max Bounds
  
  addTiles() | addProviderTiles() | addWMSTiles() #Tiles
  
  addMarkers() | addCircleMarkers() |
    addAwesomeMarkers() | addLabelOnlyMarkers() # Markers
  
  addPolylines() | addCircles() |
  addRectangles() | addPolygons() # Shapes
  
  addRasterImage(image) # Raster Data
  
  addLegend() | addLayersControl() | addControl() # Controls
  
```

.footnote[
- A Map is built by piping (`%>%`) several add* methods.

- `leaflet()`/`addXXX()` methods take an optional `options` argument for customization.
]


---

# `leaflet` Base Maps

```{r 07-leaflet-01, eval=FALSE}
leaflet() %>%
  setView(lat = 50.85045, lng = 4.34878, zoom=13) %>%
  addTiles(group="OSM") %>%
  addProviderTiles(providers$CartoDB.DarkMatter, group="Dark") %>%
  addProviderTiles(providers$CartoDB.Positron, group="Light") %>%
  addLayersControl(baseGroups=c('OSM', 'Dark','Light'))
```

---

# `leaflet` Base Maps

```{r 07-leaflet-02, echo=FALSE, fig.width=8, fig.height=6, fig.align='center'}
l <- leaflet() %>%
  setView(lat = 50.85045, lng = 4.34878, zoom=13) %>%
  addTiles(group="OSM") %>%
  addProviderTiles(providers$CartoDB.DarkMatter, group="Dark") %>%
  addProviderTiles(providers$CartoDB.Positron, group="Light") %>%
  addLayersControl(baseGroups=c('OSM','Dark','Light'))
  
frameWidget(l)
```

---

# `leaflet` Markers

- At a glance

```{r 07-leaflet-03, eval=FALSE}

leaflet(data) %>%
  
  addMarkers(
    lat = ~latitude, lon = ~longitude,
    
    options = markerOptions(),
  
    label=~label, labelOptions = labelOptions(),
    popup=~popup, popupOptions = popupOptions(),
    
    clusterOptions = clusterOptions(),
    
    group = 'Group-A')

  # Similarly 
  addCircleMarkers()  # Fixed scale Circles
  addAwesomeMarkers() # More choices for icons
  addLabelOnlyMarkers() # No icon
```

---

# `leaflet` Markers

```{r 07-leaflet-04, warning=FALSE}
quakes.df <- quakes %>% dplyr::mutate(
    mag.level = cut(mag,c(3.5,4.5,5.5,6.5),
    labels = c('> 3.5 & <=4.5', '>4.5 & <=5.5', '>5.5 & <=6.5'))) %>%
  split(.$mag.level)

l <- leaflet() %>%
  addProviderTiles(providers$Esri.OceanBasemap)

names(quakes.df) %>%
  purrr::walk( function(df) {
    l <<- l %>%
      addMarkers(data=quakes.df[[df]], lng=~long, lat=~lat,
                 label=~as.character(mag), popup=~as.character(mag),
                 group = df,
                 clusterOptions = markerClusterOptions())
  })

l <- l %>%
  addLayersControl(
    overlayGroups = names(quakes.df),
    options = layersControlOptions(collapsed = FALSE)) %>%
  addMiniMap(tiles = providers$Esri.OceanBasemap, width = 120, height=80)
```

---

# `leaflet` Markers

```{r 07-leaflet-05, echo=FALSE, fig.width=8, fig.height=6, fig.align='center'}
l %>%
frameWidget()
```

---
# `leaflet` Shapes

- At a Glance

```{r 07-leaflet-06, eval=FALSE}

leaflet(data) %>%
  addPolygons(
    label=~label, labelOptions = labelOptions(),
    popup=~popup, popupOptions = popupOptions(),
             
    # Shape Options
    options = pathOptions(),
    weight = 1, opacity=0.8, color = "#000000",
    fillColor="#ff0000", fillOpacity=0.7,
             
    # Highlighting on mouse-over
    highlightOptions = highlightOptions(
      color='#00ff00', weight = 2,
      opacity = 1, fillOpacity = 1,
      bringToFront = TRUE, sendToBack = TRUE),
             
    group = 'Group-A')

  #Similarly
  addCircles()
  addPolylines()
  addRectangles()
```

---

# Shapes Example

```{r eg leaflet-07, echo=FALSE, warning=FALSE}
library(magrittr)
fName <- system.file('extdata','world-population.geo.json',package = 'user2017.geodataviz')
spdf <- geojsonio::geojson_sp(rmapshaper::ms_simplify(readr::read_file(fName)))
spdf@data %<>% dplyr::mutate(
  AREA = as.numeric(as.character(AREA)),
  POP2005 = as.numeric(as.character(POP2005))
)

spdf <- subset(
  spdf,
  !(is.na(AREA) | AREA <1 | is.na(POP2005) | POP2005<1)
)

spdf@data %<>%
  dplyr::mutate(
  POPDENSITY = POP2005/AREA
)

spdf.world <- spdf
```

```{r 07-leaflet-08}
# spdf is a sp::SpatialPolygonsDataFrame
qpal <- colorQuantile(rev(viridis::viridis(5)),
                      spdf$POPDENSITY, n=5)

l <- leaflet(spdf, options =
               leafletOptions(attributionControl = FALSE, minzoom=1.5)) %>%
  addPolygons(
    label=~stringr::str_c(
      NAME, ' ',
      formatC(POPDENSITY, big.mark = ',', format='d')),
    labelOptions= labelOptions(direction = 'auto'),
    weight=1,color='#333333', opacity=1,
    fillColor = ~qpal(POPDENSITY), fillOpacity = 1,
    highlightOptions = highlightOptions(
      color='#000000', weight = 2,
      bringToFront = TRUE, sendToBack = TRUE)
    ) %>%
  addLegend(
    "topright", pal = qpal, values = ~POPDENSITY,
    title = htmltools::HTML("Population Density<br/>(2005)"),
    opacity = 1 )

```

---

# Shapes Example

```{r 07-leaflet-09, echo=FALSE, fig.width=8, fig.height=6, fig.align='center'}
l %>% setMapWidgetStyle() %>%
  frameWidget()
```

- Performance Tip: Use `rmapshaper::ms_simplify` with impunity!

---

# `leaflet` Other Misc. Stuff

- `addRasterImage` for adding raster image data. [Example](https://rstudio.github.io/leaflet/raster.html).

- `addMiniMap` for a small map inside the main map at a zoom offset. [Example](http://rpubs.com/bhaskarvk/leaflet-minimap).

- `addMeasure` to measure distances/area. [Example](http://rpubs.com/bhaskarvk/leaflet-measure).

- `addGraticule` adds a graticule. [Example](http://rpubs.com/bhaskarvk/leaflet-graticule).

- `addEasyButton`/`addEasyButtonBar` for customized buttons [Example](http://rpubs.com/bhaskarvk/easybutton).

- `addControl` for custom Control element.

- `addScaleBar` to display a scale.

---
# `leaflet` Projections Support

- ALL MAPS OF EARTH ARE WRONG! Obligatory [XKCD](http://xkcd.com/977/) ref.

- By Default leaflet ...

  + uses [EPSG:3857](http://epsg.io/3857) (a.k.a. Spherical Mercator) projection to display data/tiles.

  + expects tiles in EPSG:3857.

  + expects input vector/raster data in lat/lon [EPSG:4326](http://epsg.io/4326) and internally converts them to EPSG:3857 before plotting.

- Which means ...

  + You can't use tile services which provide tiles in non-spherical-Mercator projections.

  + You need to convert any vector/raster data in any non-epsg:4326 to epsg:4326 before adding to the leaflet map.

---

# `leaflet` + Proj4Leaflet

- Enter [Proj4Leaflet](https://github.com/kartena/Proj4Leaflet/) a leaflet plugin allowing use of [proj4js](https://github.com/proj4js/proj4js) to display map in non-spherical-Mercator projection.

- Basic use 

```{r 07-leaflet-10, eval=FALSE}
leaflet(options = 
          leafletOptions(crs = leafletCRS()))
```

- So now you can display data/tiles in a non-spherical-Mercator projection.

- But you still need to specify vector/raster data in EPSG:4326 (lat/lon) which will get internally converted to the custom projection specified.

- Another caveat: You can have only one projection at a time which is set during initialization. To change projection you need to destroy and rebuild the map.

---

# `leaflet` Projections Example 1.

```{r 07-leaflet-11, echo=FALSE}
spdf <- spdf.world
```

```{r 07-leaflet-12}
crs.molvidde <- leafletCRS(
  crsClass="L.Proj.CRS", code='ESRI:53009',
  proj4def= '+proj=moll +lon_0=0 +x_0=0 +y_0=0 +a=6371000 +b=6371000 +units=m +no_defs',
  resolutions = c(65536, 32768, 16384, 8192, 4096, 2048))

l <- leaflet(
  spdf,
  options = leafletOptions(
    maxZoom = 5, crs= crs.molvidde, attributionControl = FALSE)) %>%
  addGraticule(style= list(color= '#999', weight= 0.5, opacity= 1)) %>%
  addGraticule(sphere = TRUE,
               style= list(color= '#777', weight= 1, opacity= 0.25)) %>%
  addPolygons(
    label=~stringr::str_c(
      NAME, ' ', formatC(POPDENSITY, big.mark = ',', format='d')),
    labelOptions= labelOptions(direction = 'auto'),
    weight=1,color='#ffffff', opacity=1,
    fillColor = ~qpal(POPDENSITY), fillOpacity = 1,
    highlightOptions = highlightOptions(
      color='#000000', weight = 2,
      bringToFront = TRUE, sendToBack = TRUE)) 
```

---

# Projections Example 1.

```{r 07-leaflet-13, echo=FALSE, fig.width=8, fig.height=6, fig.align='center'}
l %>%
  setView(10,0,0.5) %>% setMapWidgetStyle() %>%
  frameWidget()
```

---

# Projections Example 2.

```{r 07-leaflet-14}
spdf <- rmapshaper::ms_simplify(albersusa::usa_composite())
pal <- colorNumeric(palette = "Blues", domain = spdf@data$pop_2014)

crs.laea <- leafletCRS(
  crsClass="L.Proj.CRS", code='EPSG:2163',
  proj4def='+proj=laea +lat_0=45 +lon_0=-100 +x_0=0 +y_0=0 +a=6370997 +b=6370997 +units=m +no_defs',
  resolutions = c(65536, 32768, 16384, 8192, 4096, 2048,1024, 512, 256, 128))

l <- leaflet(
  options= leafletOptions(
    worldCopyJump = FALSE, crs=crs.laea, attributionControl = FALSE)) %>%
  addPolygons(
    data=spdf, label=~stringr::str_c(
      name, ' ', formatC(pop_2014, big.mark = ',', format='d')),
    labelOptions= labelOptions(direction = 'auto'),
    weight = 1, color = "#000000",
    fillColor=~pal(pop_2014), fillOpacity=0.7,
    highlightOptions = highlightOptions(
      color='#ff0000', opacity = 1, weight = 2, fillOpacity = 1,
      bringToFront = TRUE, sendToBack = TRUE))
```

---

# Projections Example 2.

```{r 07-leaflet-15, echo=FALSE, fig.width=8, fig.height=6, fig.align='center'}
l %>%
  fitBounds(-125, 24 ,-75, 45) %>%
  setMaxBounds(-125, 24 ,-75, 45) %>%
  setMapWidgetStyle() %>% 
  frameWidget()
```

---

# `leaflet` + `tilegramsR`


```{r 07-leaflet-16, echo=FALSE}
suppressPackageStartupMessages(library(tilegramsR))

states <- FiveThirtyEightElectoralCollege.states@data$state
factpal <- colorFactor(
  colormap::colormap(nshades = length(states), 
                     colormap = colormap::colormaps$jet ), 
  states)
```

```{r 07-leaflet-17}
l <- leaflet(
  options=leafletOptions(
    crs = leafletCRS("L.CRS.Simple"),
    minZoom = -2, maxZoom = -2,
    dragging = FALSE, zoomControl = FALSE, attributionControl = FALSE)) %>%
  addPolygons(
    data=FiveThirtyEightElectoralCollege,
    weight=1,color='#000000', fillOpacity = 0.5, opacity=0.2,
    fillColor= ~factpal(state)) %>%
  addPolygons(
    data=FiveThirtyEightElectoralCollege.states, group = 'states',
    weight=2,color='#000000',
    fill = T, opacity = 1, fillOpacity = 0,
    highlightOptions = highlightOptions(weight = 4)) %>%
  addLabelOnlyMarkers(
    data=FiveThirtyEightElectoralCollege.centers,
    label = ~as.character(state),
    labelOptions = labelOptions(
      noHide = 'T', textOnly = T,
      offset=c(-8,-20), textsize = '12px'))
```

---

# Projections Example 3

```{r 07-leaflet-18, echo=FALSE, fig.width=8, fig.height=7, fig.align='center'}
l %>%
  leaflet.extras::setMapWidgetStyle() %>%
  setMapWidgetStyle() %>% 
  frameWidget()
```

---

# leaflet.extras and `leaflet.esri`

- Add-on packages for `leaflet`. Developed by yours truly to integrate the plethora of leaflet plugins available.

-  Published to CRAN just before UserR! 2017.

- Actively being developed and maintained. Contributions welcome.

- `leaflet` package will be stable and `leaflet.extras` will be very dynamic.

---

# leaflet.extras

- Add/Modify/delete/style markers/shapes using [Leaflet.Draw](http://rpubs.com/bhaskarvk/leaflet-draw).

- Add [GeoJSON](http://rpubs.com/bhaskarvk/geojsonv2), [TopoJSON](http://rpubs.com/bhaskarvk/topojsonv2), [KML](http://rpubs.com/bhaskarvk/kml), [GPX](http://rpubs.com/bhaskarvk/gpx), [CSV](http://rpubs.com/bhaskarvk/csv) files directly.
  + Customizable Markers and Shapes
  + Choropleths from polygon data w/ auto legends and bi-directional highlighting
  + Popup showing properties in a tables

- Create [Heatmap](http://rpubs.com/bhaskarvk/leaflet-heatmap) from point data.

- [Search](http://rpubs.com/bhaskarvk/leaflet-search) Markers. Geo-locate using OSM Nominatum API.

- [Pulsating](http://rpubs.com/bhaskarvk/leaflet-pulseIcon) and [Weather](http://rpubs.com/bhaskarvk/leaflet-weather) icons for markers.

- [Tiles Caching](http://rpubs.com/bhaskarvk/TileLayer-Caching), [GPS](https://github.com/stefanocudini/leaflet-gps), and many more!

---

# TopoJSON Example

```{r 07-leaflet-19}
library(leaflet.extras)
fName <- system.file('extdata','crimes_by_district.topojson', package='user2017.geodataviz')

l <- leaflet() %>%
  addBootstrapDependency() %>%
  setView(-75.14, 40, zoom = 11) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addGeoJSONChoropleth(
    readr::read_file(fName), valueProperty ='incidents',
    scale = 'OrRd', mode='q', steps = 5, padding = c(0.2,0),
    popupProperty = propstoHTMLTable(
      props = c('dist_numc', 'location', 'incidents', '_feature_id_string'),
      table.attrs = list(class='table table-striped table-bordered'),drop.na = T),
    labelProperty = JS('function(feature){return "WARD: " + feature.properties.dist_numc;}'),
    color='#ffffff', weight=1, fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      fillOpacity=1, weight=2, opacity=1, color='#000000',
      bringToFront=TRUE, sendToBack = TRUE),
    legendOptions = legendOptions(title='Crimes', position='topright'))
```

---

# TopoJSON Example

```{r 07-leaflet-20, echo=FALSE, fig.width=8, fig.height=7, fig.align='center'}
l %>%
  frameWidget()
```

---

# `crosstalk` Inter-widget communication

- Pure JavaScript solution works anywhere a normal `htmlwidget` will work.<sup>1</sup>

- Works on the concept of shared data between widgets.


```{r crosstalk-01,eval=FALSE}
library(crosstalk)
library(leaflet)
library(DT)

# Wrap data frame in SharedData
sd <- SharedData$new(quakes[sample(nrow(quakes), 100),])

# Create a filter input
filter_slider("mag", "Magnitude", sd, column=~mag, step=0.1, width=250)

# Use SharedData like a dataframe with Crosstalk-enabled widgets
bscols(
  leaflet(sd) %>% addTiles() %>% addMarkers(),
  datatable(sd, extensions="Scroller", style="bootstrap", class="compact", width="100%",
    options=list(deferRender=TRUE, scrollY=300, scroller=TRUE))
)
```

.footnote[1: Documentation: https://rstudio.github.io/crosstalk/using.html]
---

# Shiny!

- Shiny allows you to build visualizations dynamically, be they static or interactive.

- You need a Shiny server to run Shiny apps. Free/Commercial versions available from RStudio.

- Static visualizations need to destroyed and rebuilt every time the input data for the viz changes.

- Interactive visualizations (i.e. `htmlwidgets`) however can support manipulating existing viz in response to `shiny` events.

- `Plotly`, `leaflet` etc. support changing an existing map.

- `manipulateWidget` package allows building `htmlwidgets` which can respond to input events.

---

# `leaflet` + `shiny`

- Use `leafletProxy()` to update already existing map.

- Use `clear*` methods to remove stuff already on a map.

- leaflet package traps many leaflet [events](http://leafletjs.com/reference.html#map-events) and makes then available as shiny events.

- Use `observeEvent(input$<MAP_ID>_<EVENT_NAME>)` to act on these events.

---

# `leaflet` + `shiny`

```{r 07-leaflet-shiny-01, eval=FALSE}
library(shiny); library(leaflet)

r_colors <- rgb(t(col2rgb(colors()) / 255))
names(r_colors) <- colors()

ui <- fluidPage(
  leafletOutput("mymap"),
  p(),
  actionButton("recalc", "New points")
)

server <- function(input, output, session) {
  points <- eventReactive(input$recalc, {
    cbind(rnorm(40) * 2 + 13, rnorm(40) + 48)
  }, ignoreNULL = FALSE)
  output$mymap <- renderLeaflet({
    leaflet() %>%
      addProviderTiles(providers$Stamen.TonerLite, options = providerTileOptions(noWrap = TRUE)) %>%
      addMarkers(data = points())
  })
}

shinyApp(ui, server)
```

---

class: inverse middle

# Part 7: The End!


Thus marks the end of 'Geospatial visualization in R'  tutorial by Bhaskar V. Karambelkar, the first of his name!

Wait there's [more](#41).

---

# Thanks!

- So many people to thank....

  - [Bob Rudis](https://twitter.com/hrbrmstr): Mentor and teacher.
  - [Joe Cheng](https://twitter.com/jcheng): RStudio CTO and `leaflet` author.
  - [Tim Appelhans](https://twitter.com/TimSalabim3): `mapview` author and my virtual GIS guru.
  - [Kent Russels](https://twitter.com/timelyportfolio): `mapedit` and many more widgets author and collaborator for `leaflet` and `leaflet.extras`
  - [Michale Sumner](https://twitter.com/mdsumner): Author of many R GIS packages and helped me a lot understanding GIS.
  - [Kyle Walker](@kyle_e_walker): Author of `tigris`/`tidycensus`, and super helpful guy.
  - [Edzer Pebesma](https://twitter.com/edzerpebesma): R GIS super guru and author of `sf` and many many R GIS pacakges.
  - All package owners of the R GIS ecosystem.
  
